---
title: "Tutorial 2: Getting Started with R: Coding Basics"
author: "Dr. Jennifer Bigman"
format:
  html:
    toc: true
    number-sections: true
    code-copy: true
    code-overflow: wrap
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Learning objectives

By the end of this tutorial, you should be able to:

1.  Run R code from a Quarto document.
2.  Use R as a calculator.
3.  Create and inspect objects.
4.  Create vectors and perform vectorized calculations.
5.  Call functions and change their arguments.
6.  Recognize common errors involving spelling, capitalization, and punctuation.
7.  Create and inspect a simple data frame.
8.  Find help when you do not know how a function works.

## Before you begin

1.  Copy this file from the central course-hub repository into your personal tutorials-exercises repository.
2.  Open the copied `.qmd` file in RStudio.
3.  Run the code as you work through the tutorial.
4.  Save, render, commit, and push your completed copy (it will be in your personal tutorials-exercises repository).

::: callout-tip
### Useful RStudio shortcuts

| Action | Mac | Windows/Linux |
|------------------------|------------------------|------------------------|
| Run the current line or selection | `Command + Enter` | `Ctrl + Enter` |
| Insert an R code chunk | `Option + Command + I` | `Ctrl + Alt + I` |
| Insert the assignment operator `<-` | `Option + -` | `Alt + -` |
| Show all keyboard shortcuts | `Option + Shift + K` | `Alt + Shift + K` |
:::

## R, RStudio, and Quarto

These tools work together, but they are not the same thing:

- **R** is the programming language that performs calculations and analyses.
- **RStudio** is the integrated development environment, or IDE, where we write and run R code.
- **Quarto** combines text, code, results, figures, and tables in one reproducible document. We will be using Quarto for the in-class tutorials, but we will NOT be using it for the in-class exercises or out-of-class homework. This is intentional - you want experience with both.

In quarto, an R code chunk begins and ends with three backticks:

```{r}
2 + 2
```

Place your cursor inside the chunk and use the run shortcut to execute the code.

## Use R as a calculator

Run the following code one line at a time.

```{r}
2 + 2

10 - 3

6 * 4

20 / 5

2^3
```

R follows the usual order of operations.

```{r}
2 + 3 * 4

(2 + 3) * 4
```

R also includes many mathematical functions.

```{r}
sqrt(16)

log(1)

log10(100)

round(3.14159, digits = 2)
```

### Your turn 1 (will go through answers at end!)

Calculate each of the following:

1.  The mean of 12, 18, and 27.
2.  The area of a circle with radius 4.
3.  The square root of 225.
4.  (2^5).

```{r}
# Write your code here
mean(c(12,18,27))
pi * 4/2
sqrt(225)
2^5
```

## Save your code regularly after you make changes. Now is a good time to do so.

### Press Ctrl + S, click the disk button at the top, or file --> save

## Comments

R ignores any text on the same line after `#`. Comments help us understand the reasoning behind the code.

```{r}
# convert a water temperature from Celsius to Fahrenheit
12 * (9 / 5) + 32
```

Useful comments explain **why** a step was taken or describe the purpose of a section. Avoid comments that merely repeat obvious code.

```{r}
# remove a missing value before calculating the mean
fish_mass <- c(1.2, 1.5, NA, 1.8)
valid_mass <- fish_mass[!is.na(fish_mass)]
mean(valid_mass)

# do this another way
fish_mass <- c(1.2, 1.5, NA, 1.8)
mean(fish_mass, na.rm = TRUE)
```

## Objects and assignment

Most analyses involve saving values so that they can be used later. Here, we create an object with the assignment operator `<-`.

```{r}
water_temp <- 12
```

R stores the value but does not automatically print it. Type the object's name to inspect it.

```{r}
water_temp
```

Use the object in another calculation.

```{r}
water_temp_f <- water_temp * (9 / 5) + 32
water_temp_f
```

### Naming objects

Use informative names written in `snake_case`:

```{r}
mean_fish_length <- 31.5
sampling_year <- 2026
study_region <- "Delaware Bay"
```

Good names describe what an object contains.

``` text
Good: mean_fish_length
Poor:  x
Poor:  thing
Poor:  result2
```

Object names:

- must begin with a letter;
- may contain letters, numbers, `_`, and `.`;
- cannot contain spaces; and
- are case-sensitive.

```{r}
fish_length <- 30
fish_length
```

The following names refer to different objects:

```{r}
fish_length <- 30
Fish_length <- 45

fish_length
Fish_length
```

### Your turn 2

Create objects containing:

1.  Your first name.
2.  The current year.
3.  A water temperature of 14.5 °C.
4.  The same temperature converted to Fahrenheit.

Use descriptive `snake_case` names.

```{r}
# Write your code here
my_name <- "Cecelia"
current_year <- 2026
water_temp_f <- 14.5
water_temp_c <- 14.5* (9/5) + 32
```

## Vectors

A vector stores multiple values of the same basic type. Use `c()` to combine values.

```{r}
fish_lengths <- c(24, 31, 29, 35, 42)
fish_lengths
```

R applies arithmetic to every element of a numeric vector.

```{r}
fish_lengths + 1

fish_lengths * 2

fish_lengths / 10
```

This behavior is called **vectorization**.

You can subset vectors.

```{r}

	animals <- c("mouse", "rat", "dog", "cat")

	animals[2] # return second element
	
	animals[c(3, 2)] # return second and third elements

	
	weight_g <- c(21, 34, 39, 54, 55)
	
	weight_g > 50 # return logicals with TRUE for the indices that are > 50
	
	weight_g[weight_g > 50] # select values > 50
	
```


### Common summary functions

```{r}
length(fish_lengths)

sum(fish_lengths)

mean(fish_lengths)

min(fish_lengths)

max(fish_lengths)

range(fish_lengths)
```

### Logical comparisons

Comparisons return `TRUE` or `FALSE`.

```{r}
fish_lengths > 30

fish_lengths == 29

fish_lengths != 29
```

Common comparison operators are:

| Operator | Meaning                  |
|----------|--------------------------|
| `>`      | greater than             |
| `<`      | less than                |
| `>=`     | greater than or equal to |
| `<=`     | less than or equal to    |
| `==`     | equal to                 |
| `!=`     | not equal to             |

::: callout-warning
Use `==` to test equality. A single `=` is generally used to provide named arguments inside functions.
:::

Count how many fish are longer than 30 cm.

```{r}
sum(fish_lengths > 30)
```

Subset a vector using square brackets.

```{r}
fish_lengths[fish_lengths > 30]
```

### Your turn 3

Create this vector of water temperatures:

```{r}
temperatures <- c(8.2, 10.5, 12.1, 9.8, 14.3, 11.6)
```

Then calculate:

1.  The number of observations.
2.  The mean temperature.
3.  The minimum and maximum temperatures.
4.  How many temperatures are greater than 11 °C.
5.  A vector containing only temperatures greater than 11 °C.

```{r}
# Write your code here
length(temperatures)
mean(temperatures)
min(temperatures)
max(temperatures)
sum(temperatures>11)
temperatures[temperatures>11]
```

## Functions and arguments

Functions perform actions. A function call generally has this structure:

``` text
function_name(argument1 = value1, argument2 = value2)
```

For example, `round()` takes a number and an argument controlling the number of decimal places.

```{r}
round(3.14159, digits = 2)

round(3.14159, digits = 4)
```

Some arguments have default values. You only need to supply an argument when you want to change the default behavior.

```{r}
seq(from = 1, to = 10)

seq(from = 1, to = 10, by = 2)
```

You can often omit the names of the first arguments:

```{r}
seq(1, 10, by = 2)
```

Using argument names is often clearer for beginners and for less familiar functions.

### Missing values

R represents a missing value with `NA`.

```{r}
fish_mass <- c(1.2, 1.5, NA, 1.8)
fish_mass
```

A calculation involving an unknown value is usually unknown.

```{r}
mean(fish_mass)
```

Many summary functions include an `na.rm` argument. Set it to `TRUE` to remove missing values during the calculation.

```{r}
mean(fish_mass, na.rm = TRUE)
```

Check which values are missing:

```{r}
is.na(fish_mass)
```

Count the missing values:

```{r}
sum(is.na(fish_mass))
```

### Your turn 4

Use the vector below.

```{r}
survey_catch <- c(12, 8, NA, 15, 21, NA, 10)
```

Calculate:

1.  The mean while ignoring missing values.
2.  The total while ignoring missing values.
3.  The number of missing values.
4.  The number of observed values.

```{r}
# Write your code here
mean(survey_catch, na.rm = TRUE)
sum(survey_catch, na.rm = TRUE)
sum(is.na(survey_catch))
length(survey_catch) - sum(is.na(survey_catch))
```

## Character and logical vectors

Vectors can store text.

```{r}
species <- c("cod", "haddock", "hake", "flounder")
species
```

Text must be enclosed in quotation marks.

Vectors can also store logical values.

```{r}
is_large <- c(FALSE, TRUE, FALSE, TRUE)
is_large
```

A single vector should generally contain one basic type. Mixing types may cause R to convert values to a common type.

```{r}
mixed_values <- c(1, 2, "three")
mixed_values
```

## Data frames

A data frame stores observations in rows and variables in columns.

Create a small fisheries data frame:

```{r}
fish_data <- data.frame(
  species = c("cod", "cod", "haddock", "haddock", "hake"),
  length_cm = c(42, 38, 31, 35, 29),
  region = c("north", "south", "north", "south", "south")
)

fish_data
```

Inspect its structure.

```{r}
str(fish_data)
```

View the first rows.

```{r}
head(fish_data)
```

Get a summary of each column.

```{r}
summary(fish_data)
```

Use `$` to extract one column.

```{r}
fish_data$length_cm
```

Calculate the mean fish length.

```{r}
mean(fish_data$length_cm)
```

Use a logical condition to select rows.

```{r}
fish_data[fish_data$length_cm > 35, ]
```

The comma matters:

``` text
data_frame[rows, columns]
```

Leaving the column position blank means “keep all columns.”

### Your turn 5

Using `fish_data`:

1.  Find the maximum fish length.
2.  Count the number of fish from the south region.
3.  Display only fish that are at least 35 cm long.
4.  Display only cod observations.

```{r}
# Write your code here
max(fish_data$length_cm)
sum(fish_data$region == "south")
fish_data[fish_data$length_cm >= 35, ]
fish_data[fish_data$species == "cod", ]
```


### Combining conditions when selecting rows 

So far, we have selected rows using one condition at a time. We can also
combine multiple logical conditions.

First, create a slightly more complicated data frame containing a missing
fish length and a logical column indicating whether each fish was caught.

```{r}
survey_data <- data.frame(
  species = c(
    "cod",
    "cod",
    "haddock",
    "hake",
    "haddock",
    "hake",
    "cod"),
  length_cm = c(42, 38, NA, 31, 35, 29, 44),
  region = c(
    "north",
    "south",
    "north",
    "south",
    "south",
    "north",
    "north"),
  caught = c(
    TRUE,
    TRUE,
    FALSE,
    TRUE,
    TRUE,
    FALSE,
    TRUE))

survey_data
```

### Select rows using a character condition

The following condition produces one `TRUE` or `FALSE` value for each row.

```{r}
cod_rows <- survey_data$species == "cod"

cod_rows
```

We can use that logical vector to select rows from the data frame.

```{r}
cod_observations <- survey_data[
  cod_rows,
]

cod_observations
```

Recall that data-frame subsetting follows this structure:

```text
data_frame[rows, columns]
```

The comma is important. Leaving the space after the comma blank means that
R should retain all columns.

### Select rows using an existing logical column

The `caught` column already contains `TRUE` and `FALSE` values.

```{r}
survey_data$caught
```

Therefore, we can use it directly to select rows. R retains rows where
`caught` is `TRUE`.

```{r}
caught_observations <- survey_data[
  survey_data$caught,
]

caught_observations
```

This is equivalent to writing:

```{r}
caught_observations <- survey_data[
  survey_data$caught == TRUE,
]

caught_observations
```

The shorter version is usually preferred because `survey_data$caught` is
already a logical vector.

### Select rows using a calculated value

First, calculate the mean of the observed fish lengths.

```{r}
mean_observed_length <- mean(
  survey_data$length_cm,
  na.rm = TRUE
)

mean_observed_length
```

Next, identify fish with lengths greater than the mean.

```{r}
survey_data$length_cm > mean_observed_length
```

Notice that the fish with a missing length produces `NA` rather than
`TRUE` or `FALSE`. We can use `!is.na()` to identify rows with observed
lengths.

```{r}
!is.na(survey_data$length_cm)
```

The `!` symbol means “not.” Therefore:

```r
!is.na(survey_data$length_cm)
```

is `TRUE` when the length is not missing.

Use `&` to require both conditions to be true:

```{r}
above_mean_rows <-
  !is.na(survey_data$length_cm) &
  survey_data$length_cm > mean_observed_length

above_mean_rows
```

Now use the logical vector to select the appropriate rows.

```{r}
above_mean_observations <- survey_data[
  above_mean_rows,
]

above_mean_observations
```

### Calculate a summary for selected observations

We can first select the lengths of fish that were caught:

```{r}
caught_lengths <- survey_data$length_cm[
  survey_data$caught
]

caught_lengths
```

Then calculate their mean:

```{r}
mean(
  caught_lengths,
  na.rm = TRUE
)
```

### Your turn 6

Using `survey_data`:

1. Create a data frame containing only observations from the north region.
2. Create a data frame containing only observations with nonmissing lengths.
3. Create a data frame containing only cod that were caught.
4. Calculate the mean length of cod, ignoring missing values.

```{r}
# Write your code here
```

## Reading errors

Errors are a normal part of programming. Read the message carefully and inspect the code immediately above it.

### Object not found

```{r}
#| error: true

fish_weight <- 3.2
fish_Weight
```

R is case-sensitive: `fish_weight` and `fish_Weight` are different names.

### Function not found

```{r}
#| error: true

squareroot(16)
```

The correct function is `sqrt()`.

### Incomplete expression

A missing quotation mark, comma, or parenthesis may cause R to wait for more input. In the Console, a `+` prompt usually means the expression is incomplete. Press `Esc` to cancel it, then correct the code.

### Debugging checklist

When code does not work, check:

1.  Is every object and function spelled correctly?
2.  Does capitalization match?
3.  Are quotation marks and parentheses paired?
4.  Are commas in the correct places?
5.  Did you run the code that creates the object?
6.  Does the error identify a particular line or object?

## Getting help

Open the help page for a function with `?`.

```{r}
?mean
```

You can also use `help()`.

```{r}
help(mean)
```

Inspect the arguments of a function.

```{r}
args(mean)
```

Try tab completion in RStudio:

1.  Type the first few letters of an object or function.
2.  Press `Tab`.
3.  Select from the available completions.


## Final challenge

Create a data frame named `survey_data` with the following columns:

| station | species | length_cm | caught |
|---------|---------|----------:|--------|
| A       | cod     |        41 | TRUE   |
| B       | haddock |        33 | TRUE   |
| C       | cod     |        38 | FALSE  |
| D       | hake    |        27 | TRUE   |
| E       | haddock |        36 | TRUE   |

Then use R to answer these questions:

1.  What is the mean length?
2.  How many fish were caught?
3.  Which observations are cod?
4.  Which observations are longer than the mean length?
5.  What is the mean length of the fish that were caught?

```{r}
# Create survey_data and complete the analysis here
```



## Save and render your work

Before leaving:

1.  Save the `.qmd` file.
2.  Click **Render**.
3.  Confirm that the HTML document is created without errors.
4.  Review the rendered document.
5.  Commit your changes with a meaningful message.
6.  Push the commit to GitHub.

Example commit message:

``` text
Complete introductory R tutorial
```

## Key ideas

- R code must be precise: spelling, capitalization, and punctuation matter.
- Use `<-` to assign a value to an object.
- Use descriptive `snake_case` object names.
- Vectors store multiple values and support vectorized operations.
- Functions perform actions and arguments control how they behave.
- Missing values are represented by `NA`.
- Data frames organize observations in rows and variables in columns.
- Errors are information that can help you locate and fix a problem.
- Save code in a script or Quarto document so your work is reproducible.

## Sources and further reading

This tutorial was developed using ideas and examples adapted into an original course format from:

- Wickham, H., Çetinkaya-Rundel, M., and Grolemund, G. *R for Data Science, 2nd edition*: [Workflow basics](https://r4ds.hadley.nz/workflow-basics.html)
- Zuur, A. and Ieno, E. *An Introduction to R*: [Getting started](https://intro2r.com/getting-started.html)
- Posit: [Quarto documentation](https://quarto.org/docs/guide/)
- R Core Team: [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html)
